---
title: "The Leslie Matrix - revisited"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

A population of fish we're interested in has four life stages: eggs (E), fry (F), juvenile (J), breeding adult (A). You are told that annually:

- Each breeding adult will produce, on average, 600 eggs
- 5% of eggs will survive to become fry
- 20% of fry will survive to become juveniles
- 16% of juveniles will survive to become adults
- Adult survival rate year to year is 90%

Recall the Leslie matrix from EDS 212. Matrix multiplication (using the dot product `%*%` of the matrix with the populations at each life stage) yields the population structure at the following year. 

### Create the Leslie matrix:

```{r}
# Make the Leslie Matrix:
fish_leslie <- matrix(c(0, 0, 0, 600, 0.05, 0, 0, 0, 0, 0.20, 0, 0, 0, 0, 0.16, 0.9), nrow = 4, ncol = 4, byrow = TRUE)

# Check it out: 
fish_leslie
```

The initial population structure is given by: 
```{r}
# Initial population structure (Roe, Fry, Juvenile, Adult): 
fish_year0 <- c(5000, 8000, 600, 400)
```


Add line-by-line comments wheres prompted in the code below, explaining what each is doing. Make sure that you are talking with your neighbors about what is really being created at each step. 

```{r}
# Model projections

# ------ WHAT HAPPENS HERE? ------ #
# ------ comment line-by-line ---- #

# The line below...
time_yr <- seq(from = 0, to = 8, by = 1)
#time_yr is a vector from 0-8, by 1, species is reproducing annually over 8 years, so creating 9 values

# The line below...
proj_year <- matrix(ncol = length(fish_year0), nrow = length(time_yr))
#number of columns is 4, number of rows is 9, this is an empty matrix waiting to be filled

# The line below...
proj_year[1, ] <- fish_year0
# in row 1, and considering all the columns in row 1 = fish_year0
# [rows, col]

# The for loop below...
for (i in 2:length(time_yr)) {
  proj_year[i,] <- fish_leslie %*% proj_year[i-1,]
}
# start at index 2 because index 1 was given as initial conditions, this will run 8 times, from years 2-9
# take the dot product of fish leslie matrix and proj year to create each year from previous time step over every column for each row
# make sure to use the %*% rather than just *
# this can also be written with 1:length(time_yr), proj_year[i+1,], proj_year[i,] replacing those indexes in the for loop

# you can also flip the rows and col of the matrix to have rows = life stages and col = years, and then you have to flip [i,] --> [,i] and same for switching [i-1,] --> [,i-1]

# The rest is wrangling & visualization (run to check it out):
colnames(proj_year) <- c("eggs", "fry", "juvenile", "adult")

proj_df <- data.frame(time_yr, proj_year) %>% 
  pivot_longer(cols = -time_yr, names_to = "lifestage", values_to = "stage_population")

ggplot(data = proj_df, aes(x = time_yr, y = stage_population)) +
  geom_line(aes(color = lifestage)) +
  scale_y_log10()

```

Test different combinations of survival / reproduction and starting population structures to see what happens! 
